Some examples in extended logic counting classes

### 2016-12-19 LICS/appendix.tex
19 dic

>>>>

For condition (1) in \ref{cap-def}, let $\alpha \in \eqso(\fo)$ be defined over a relational signature $\R$. Let $\ell$ be the number of algebraic operators in $\alpha$. We will prove by induction over $\ell$ that there is a $\np$ machine $M$ such that $\alpha(\enc(\A))$ is equal to the number of accepting paths of $M$ on input $\enc(\A)$ for every $\A\in\ostr[\R]$. Let $\ell = 0$. Then $\alpha$ is either a formula in $\fo$ or a constant, and both functions can be easily computed in polynomial time. Suppose the hypothesis holds for $\ell$ from 1 through $n-1$. Let $\alpha$ have $n$ algebraic operators and let $\A$ be an ordered structure over $\R$ with domain $A$. If $\alpha = (\beta \add \gamma)$, then there exists a $\np$ machine for both $\beta$ and $\gamma$. Consider a machine that chooses between 0 or 1, if it is 0 (1), it calculates the function for $\beta$ ($\gamma$). The number of accepting paths in this $\np$ machine is equal to the function for $(\beta \add \gamma)$. If $\alpha = (\beta \mult \gamma)$, consider a machine that evaluates $\beta$, and in each accepting path, it continues evaluating $\gamma$. The number of accepting paths in this $\np$ machine is equal to the function for $(\beta \mult \gamma)$. If $\alpha = \sa{x}\beta$, consider a machine that chooses some $a\in A$ and then evaluates $\beta$ over $a$. If $\alpha = \pa{x}\beta$, consider a machine that evaluates $\beta$ consecutively over every element in $A$. If $\alpha = \sa{X}\beta$, consider a machine that guesses $P \subseteq A^{arity(X)}$ and then evaluates $\beta$ over $P$. This covers all possible cases for $\alpha$. Let $f$ be the function that this machine computes. We have that $f$ is a $\shp$ function over $\R$ and that for every $\A\in\ostr[\R]$: $f(\enc(\A)) = \alpha(\enc(\A))$.

For condition (2) in \ref{cap-def}, note that Saluja et al. \cite{DBLP:journals/jcss/SalujaST95} proved that $\shp = \#\fo$. We also have that $\sqso(\fo)$ captures $\#\fo$ over ordered structures so for each $f\in \shp$ let $\alpha \in \sqso$ be its corresponding formula. Since this formula is also in $\eqso$, the condition holds.

### 2016-12-19 LICS/counting.tex
19 dic

Assume that $\alpha$ is a sentence in $\qso$. Then slightly abusing notation, we also use $\alpha$ to denote a function defined over $\R$  such that for every $\A \in \ostr[\R] $:
\begin{eqnarray*}
	\alpha(\enc(\A)) & = & \sem{\alpha}(\A).
\end{eqnarray*}
\begin{definition}
	Let $\FF$ a fragment of $\qso$ and $\CC$ a function complexity class. Then $\FF$ {\em captures} $\CC$ if:
	\begin{itemize}
		\item for every $\alpha$ in $\FF$, let $\R$ be some signature in which $\alpha$ is in. There exists $f \in \CC$ such that for every 
		$\A \in \ostr[\R]$:  $\alpha(\enc(\A)) = f(\enc(\A))$; and

		\item for every $f \in \CC$, let $f$ be defined over $\R$. There exists $\alpha$ in $\FF$ such that for every 
		$\A \in \ostr[\R]$:  $f(\enc(\A)) = \alpha(\enc(\A))$.
	\end{itemize}
\end{definition}
%In the previous definition, if $\KK$ is the class of all finite structures, then we say that $\FF$ captures $\CC$ over the class of all structures, and if $\KK$is the class of all ordered finite structures, then we say that $\FF$ captures $\CC$ over the class of ordered structures.

\marcelo{Hacer una pasada a la demostracion del siguiente teorema considerando que solo usamos estructuras ordenadas, y teniendo en cuenta que las funciones estan definidas sobre relational signatures.}
\martin{done}
\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over the class of ordered structures.
\end{theorem}

\marcelo{Hacer una pasada a la demostracion del siguiente teorema basada en la demostracion para LFP.}
\martin{done.}
\begin{theorem} \label{qfo-pfp-cap}
	$\qfo(\pfp)$ captures $\nfpspace$ over the class of ordered structures.
\end{theorem}

\begin{theorem} \label{eqso-fo-cap}
	$\eqso(\fo)$ captures $\shp$ over the class of ordered structures.
\end{theorem}

\begin{theorem} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over the class of ordered structures.
\end{theorem}

>>>>

Assume that $\alpha$ is a sentence in $\qso$. Then slightly abusing notation, we also use $\alpha$ to denote a function defined over $\R$ such that for every $\A \in \ostr[\R] $:
\begin{eqnarray*}
	\alpha(\enc(\A)) & = & \sem{\alpha}(\A).
\end{eqnarray*}
\begin{definition} \label{cap-def}
	Let $\FF$ a fragment of $\qso$ and $\CC$ a function complexity class. Then $\FF$ {\em captures $\CC$ over ordered structures} if

	\begin{itemize}
	\item[(1)] for every $\alpha$ in $\FF$ over a signature $\R$. There exists $f \in \CC$ such that for every $\A \in \ostr[\R]$:  $\alpha(\enc(\A)) = f(\enc(\A))$; and

	\item[(2)] for every $f \in \CC$, let $f$ be defined over $\R$. There exists $\alpha$ in $\FF$ such that for every $\A \in \ostr[\R]$:  $f(\enc(\A)) = \alpha(\enc(\A))$.
	\end{itemize} 
\end{definition}
%In the previous definition, if $\KK$ is the class of all finite structures, then we say that $\FF$ captures $\CC$ over the class of all structures, and if $\KK$is the class of all ordered finite structures, then we say that $\FF$ captures $\CC$ over the class of ordered structures.

\marcelo{Hacer una pasada a la demostracion del siguiente teorema considerando que solo usamos estructuras ordenadas, y teniendo en cuenta que las funciones estan definidas sobre relational signatures.}
\martin{done}
\begin{theorem} \label{captfp}
	$\qfo(\lfp)$ captures $\fp$ over ordered structures.
\end{theorem}

\begin{theorem} \label{qfo-pfp-cap}
	$\qfo(\pfp)$ captures $\nfpspace$ over ordered structures.
\end{theorem}

Saluja et. al \cite{DBLP:journals/jcss/SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\fo$. For a formula $\varphi(x,X)$ over $\R$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\ostr[\R]$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL)$ captures $\#\LL$ over ordered structures and that $\sqso(\LL) \subseteq \eqso(\LL)$.

\begin{theorem} \label{eqso-fo-cap}
	$\eqso(\fo)$ captures $\shp$ over ordered structures.
\end{theorem}

\begin{theorem} \label{eqso-eso-cap}
	$\eqso(\eso)$ captures $\spp$ over ordered structures.
\end{theorem}

\begin{theorem} \label{qso-pfp-cap}
	$\qso(\pfp)$ captures $\fpspace$ over ordered structures.
\end{theorem}

### 2016-12-19 LICS/framework.tex
19 dic

Consider $\bbN$ as the set of non-negative integers. Then given a relational signature $\R$, the set of quantitative second-order logic formulas (or just $\qso$-formulas) over $\R$ is given by the following grammar:
\begin{multline}
\label{eq-def-qso}
\alpha := \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \\ \pa{x} \alpha \ \mid \ \sa{X} \alpha \ \mid \ \pa{X} \alpha 
\end{multline}
where $\varphi$ is an $\so$-formula over $\R$, $s \in \bbN$, $x \in \fv$ and $X \in \sv$. If $\R$ is not mentioned, $\qso$ refers to the union of the sets of $\qso$ formulas over $\R$, for every relational signature $\R$.

Let $\R$ be a relational signature, $\A$ be an ordered finite $\R$-structure with domain $A$, $v$ a first-order assignment for $\A$ and $V$ a second-order assignment for $\A$. Then the \emph{evaluation} of a $\qso$-formula $\alpha$ over $(\A, v, V)$ is defined as a function $\sem{\alpha}$ that on input $(\A, v, V)$ returns a non-negative integer in $\bbN$. Formally, the function $\sem{\alpha}$ is recursively defined as follows:
$$
\renewcommand{\arraystretch}{1.7}
\begin{array}{rcl} 
\sem{\varphi}(\A, v, V) & = & 
\begin{cases}
1 & \mbox{if } (\A, v, V) \models \varphi \\
0 & \mbox{otherwise}
\end{cases}\\
\sem{s}(\A, v, V) & = & s \\
\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
\sem{\sa{x} \alpha}(\A, v, V) & = & \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\pa{x} \alpha}(\A, v, V) & = & \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\sa{X} \alpha}(\A, v, V) & = & \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
\sem{\pa{X} \alpha}(\A, v, V) & = & \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
\end{array}
$$
A formula in $\qso$ can mention the usual quantifiers in $\so$ (that is, $\exists x$ and $\exists X$) and the quantifiers that make use of addition and multiplication (that is, $\Sigma x$, $\Pi x$, $\Sigma X$ and $\Pi X$), which are called {\em quantitative quantifiers} . A $\qso$-formula $\alpha$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\alpha$ is under the scope of a usual quantifier or a quantitative quantifier. It is important to notice that if $\alpha$ is a $\qso$-sentence over a relational signature $\R$, then for every ordered finite $\R$-structure $\A$, first-order assignments $v_1$, $v_2$ for $\A$ and second-order assignments $V_1$, $V_2$ for $\A$, it holds that:
\begin{eqnarray*}
	\sem{\alpha}(\A, v_1, V_1) & = & \sem{\alpha}(\A, v_2, V_2).
\end{eqnarray*}
Thus, in such a case we use the term $\sem{\alpha}(\A)$ to denote $\sem{\alpha}(\A, v, V)$, for some arbitrary first-order assignment $v$ for $\A$ and some arbitrary second-order assignment $V$ for $\A$.

>>>>

Given a relational signature $\R$, the set of quantitative second-order logic formulas (or just $\qso$-formulas) over $\R$ is given by the following grammar:
\begin{multline}
\label{eq-def-qso}
\alpha := \varphi \ \mid \ s \ \mid \ (\alpha \add \alpha) \ \mid\ (\alpha \mult \alpha) \ \mid \ \sa{x} \alpha \ \mid \\ \pa{x} \alpha \ \mid \ \sa{X} \alpha \ \mid \ \pa{X} \alpha 
\end{multline}
where $\varphi$ is an $\so$-formula over $\R$, $s \in \bbN$, $x \in \fv$ and $X \in \sv$. If $\R$ is not mentioned, $\qso$ refers to the union of the sets of $\qso$ formulas over $\R$, for every relational signature $\R$.

Let $\R$ be a relational signature, $\A$ be an ordered finite $\R$-structure with domain $A$, $v$ a first-order assignment for $\A$ and $V$ a second-order assignment for $\A$. Then the \emph{evaluation} of a $\qso$-formula $\alpha$ over $(\A, v, V)$ is defined as a function $\sem{\alpha}$ that on input $(\A, v, V)$ returns a number in $\bbN$. Formally, the function $\sem{\alpha}$ is recursively defined as follows:
$$
\renewcommand{\arraystretch}{1.7}
\begin{array}{rcl} 
\sem{\varphi}(\A, v, V) & = & 
\begin{cases}
1 & \mbox{if } (\A, v, V) \models \varphi \\
0 & \mbox{otherwise}
\end{cases}\\
\sem{s}(\A, v, V) & = & s \\
\sem{\alpha_1 \add \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) + \sem{\alpha_2}(\A, v, V)\\
\sem{\alpha_1 \mult \alpha_2}(\A, v, V) & = & \sem{\alpha_1}(\A, v, V) \cdot \sem{\alpha_2}(\A, v, V)\\ 
\sem{\sa{x} \alpha}(\A, v, V) & = & \displaystyle \sum_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\pa{x} \alpha}(\A, v, V) & = & \displaystyle \prod_{a \in A} \sem{\alpha}(\A,v[a/x],V)\\
\sem{\sa{X} \alpha}(\A, v, V) & = & \displaystyle \sum_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])\\
\sem{\pa{X} \alpha}(\A, v, V) & = & \displaystyle \prod_{B \subseteq A^{\arity(X)}} \sem{\alpha}(\A, v, V[B/X])
\end{array}
$$
A formula in $\qso$ can mention the usual quantifiers in $\so$ (that is, $\exists x$ and $\exists X$) and the quantifiers that make use of addition and multiplication (that is, $\Sigma x$, $\Pi x$, $\Sigma X$ and $\Pi X$), which are called {\em quantitative quantifiers} . A $\qso$-formula $\alpha$ is said to be a \emph{sentence} if it does not have any free variable, that is, every variable in $\alpha$ is under the scope of a usual quantifier or a quantitative quantifier. It is important to notice that if $\alpha$ is a $\qso$-sentence over a relational signature $\R$, then for every ordered finite $\R$-structure $\A$, first-order assignments $v_1$, $v_2$ for $\A$ and second-order assignments $V_1$, $V_2$ for $\A$, it holds that:
\begin{eqnarray*}
	\sem{\alpha}(\A, v_1, V_1) & = & \sem{\alpha}(\A, v_2, V_2).
\end{eqnarray*}
Thus, in such a case we use the term $\sem{\alpha}(\A)$ to denote $\sem{\alpha}(\A, v, V)$, for some arbitrary first-order assignment $v$ for $\A$ and some arbitrary second-order assignment $V$ for $\A$.

### 2016-12-19 LICS/preliminaries.tex
19 dic

\marcelo{Dado que solo vamos a considerar estructuras ordenadas, hay que revisar y simplificar la notacion. Mantenemos las notaciones $\str[\R]$ y $\ostr[\R]$.}

A $\mathbb{N}$ relational signature $\R$ is a finite set $\{R_1, \ldots, R_k\}$, where each $R_i$ ($1 \leq i \leq k$) is a relation name with an associated arity greater than 0, which is denoted by $\arity(R_i)$. A finite structure over $\R$ (or just finite $\R$-structure) is a tuple $\A = \langle A, R_1^\A, \ldots, R_k^\A \rangle$ such that $A$ is a finite set and $R_i^\A \subseteq A^{\arity(R_i)}$ for every $i \in \{1, \ldots, k\}$. A finite $\R$-structure $\A$ is said to be ordered if $<$ is a binary predicate name in $\R$ and $<^\A$ is a linear order on $A$. Let $\str[\R]$ be the class of all finite $\R$-structures and $\ostr[\R]$ be the class of all ordered finite $\R$-structures.

@@

\marcelo{Agregar las definiciones de LFP y PFP}
\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements, and recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a relational signature $\R$,  function $f$ is said to be defined over $\R$ if $f$ is a total function from $\{\enc(\A) \mid \A \in \ostr[\R]\}$ to $\nat$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class} if every $f \in \CC$ is a function defined over some relational signature $\R$.

>>>>

A relational signature $\R$ (or just signature) is a finite set $\{R_1, \ldots, R_k\}$, where each $R_i$ ($1 \leq i \leq k$) is a relation name with an associated arity greater than 0, which is denoted by $\arity(R_i)$. A finite structure over $\R$ (or just finite $\R$-structure) is a tuple $\A = \langle A, R_1^\A, \ldots, R_k^\A \rangle$ such that $A$ is a finite set and $R_i^\A \subseteq A^{\arity(R_i)}$ for every $i \in \{1, \ldots, k\}$. A finite $\R$-structure $\A$ is said to be ordered if $<$ is a binary predicate name in $\R$ and $<^\A$ is a linear order on $A$. Let $\str[\R]$ be the class of all finite $\R$-structures and $\ostr[\R]$ be the class of all ordered finite $\R$-structures.

@@

\marcelo{Agregar las definiciones de LFP y PFP}
\subsection{Function complexity classes}
Assume that $\R = \{R_1, \ldots, R_k\}$ is a relational signature and $\A$ is an ordered finite $\R$-structure with a domain $A$ containing $n$ elements. Recall that  $<$ is a linear order on $A$, say $a_1 < a_2 < \ldots < a_n$. For every $i \in \{1, \ldots, k\}$, define the encoding of $R_i^\A$, denoted by $\enc(R_i^\A)$, as the following binary string. Assume that $\ell = \arity(R_i)$ and consider an enumeration of the $\ell$-tuples over $A$ in the lexicographic order induced by $<$ (that is, $(a_1, \ldots, a_1, a_1)$, $(a_1, \ldots, a_1, a_2)$, $\ldots$, $(a_n, \ldots, a_n, a_{n-1})$, $(a_n, \ldots, a_n, a_n)$). Then let $\enc(R_i^\A)$ be a binary string of length $n^\ell$ such that the $i$-th bit of $\enc(R_i^\A)$ is 1 if the $i$-th tuple in the previous enumeration belongs to $R_i^\A$, and 0 otherwise. Moreover, define the encoding of $\A$, denoted by $\enc(\A)$, as the following binary string~\cite{L04}:
\begin{eqnarray*}
	\enc(\A) & = & 0^n \, 1 \, \enc(R_1^\A) \, \cdots \, \enc(R_k^\A).
\end{eqnarray*}
Given a relational signature $\R$,  function $f$ is said to be {\em defined over $\R$} (or just {\em over $\R$}) if $f$ is a total function from $\{\enc(\A) \mid \A \in \ostr[\R]\}$ to $\nat$. Moreover, a set $\CC$ of functions is said to be a {\em function complexity class} if every $f \in \CC$ is a function defined over some relational signature $\R$.

### 2016-12-19 LICS/saluja.tex
19 dic

Saluja et. al \cite{DBLP:journals/jcss/SalujaST95} define a family of counting classes $\#\LL$ for each fragment $\LL$ of $\fo$. For a formula $\varphi(x,X)$, the function $f_{\varphi(x,X)}$ is defined as
\[
f_{\varphi(x,X)}(\A) = \vert \{\langle e,P\rangle\mid \A\models\varphi(e,P)\}\vert.
\]
for each $\A\in\str$. A function $f:\Sigma^*\to\nat$ is in $\#\LL$ if there exists an $\LL$ formula $\varphi(x,X)$ such that $f = f_{\varphi(x,X)}$. Note that for each fragment $\LL$ of $\fo$, we have that $\#\LL \subseteq \eqso(\LL)$. In fact we need a more restricted fragment of $\eqso$. Let $\sqso(\LL)$ be a fragment of $\qso$ defined by the following grammar:
\[
\alpha := \sa{x}\varphi \ \mid \ \sa{X} \varphi \ \mid \ \sa{x}\alpha \ \mid \ \sa{X}\alpha,
\]
where $\varphi$ is an $\LL$-formula. It can be seen that $\sqso(\LL) = \#\LL$. We are interested in, for each $\fo$-fragment $\LL$, the biggest fragment of $\eqso$ that is contained in $\#\LL$.

>>>>

We are interested in, for each $\fo$-fragment $\LL$, the biggest fragment of $\eqso$ that is contained in $\#\LL$.
